Supplementary Information

### 2021-01-13 doc/manuscript/MEE_manuscript.Rmd
writing/ refs related to issue #109

**Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [**REFS**; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011]. 
However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle.  
A variety of methods have been used to analyze directional changes in grwoth rate through time, including "conservative detrending", where a mathematical function () is fit to growth trends in each individual core and residuals analyzed for trends and "basal area correction", which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
None of these methods are fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014; @peters_detecting_2015], leaving great uncertainty as to what the tree-ring record can tell us about the impact of slowly changing environmental drivers such as rising CO~2~ [@sullivan_effect_2016; @walker_integrating_2020]. 
To disentangle $DBH$ from time and detect directional grwoth trends, it is necessary to simultaneously account for the influences of changing tree size and environmental drivers, including climate.

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
We note that a simple modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could be used to account for directional changes in climate sensitivity, which have commonly been observed [REFS- Maxwell, Helcoski; @peltier_tree_2020; @zuidema_recent_2020].

@@

Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations that could be analyzed (Fig. `r year_fig`).
This finding was consistent with an independent analysis (Appendix `r decadal_analysis_appendix`, Figs. `r decadal_analysis_figs_all`) and corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
*(have there been analyses of this at any other sites?)*
Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015; @brienen_tree_2017], decent coverage of $DBH$ x $year$ matrices for the species analyzed, together with general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs. `r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is relatively reliable, *albiet* still subject to the influence of some potential sampling and survivorship biases [@brienen_detecting_2012; @brienen_tree_2017], evaluation of which is beyond the scope of the current analysis.
At a minimum, it provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig. `r size`).
For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
Indeed, in a comparative analysis of methods to detect growth trends, @peters_detecting_2015 found that approach similar to ours ("regional curve standardization") to be among the most robust and recommended that future analyses take an approach very similar to that used here.

@@

Changes over time in growth rate are attributable to directionally changing environmental conditions, consisting of some combination of changes in the biotic neighborhood (e.g., competition, crown position) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
The majority of negative growth trends observed here are probably attributable to increasing competition as secondary stands age. 
Several of our sites (**HKK, SCBI, LDW, HF, ...** *Coauthors, I need help here.*) are known to have suffered major disturbances within or shortly prior to the analysis time frame [Appendix `r site_details_appendix`; REFS: @baker_disturbance_2005; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
*Indeed, stand succession has been recognized as a driver of growth declines in tree ring records... (REFS).*
Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.
For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.

>>>>

**Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [**REFS**; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011]. 
However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [REFS; @cherubini_potential_1998; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @hember_tree_2019].  
A variety of methods have been used to analyze directional changes in growth rate through time, including conservative detrending, where a mathematical function (e.g., negative exponential or stiff spline) is fit to growth trends in each individual core and residuals analyzed for trends and basal area correction, which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
A more reliable alternative is regional curve standardization, which uses an average age or size curve to describe ontogeny (*Esper et al. 2003*) and can be used in the context of multiple regressions or mixed-effects models [*REFS in @hember_tree_2019*].
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been observed [REFS- Maxwell, Helcoski; @peltier_tree_2020; @zuidema_recent_2020].

@@

Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig. `r year_fig`).
This finding was consistent with an independent analysis of growth-$DBH$ trends by decade (Appendix `r decadal_analysis_appendix`, Figs. `r decadal_analysis_figs_all`) and corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015; @brienen_tree_2017], general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs. `r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is robust in its analytical structure, *albiet* still subject to the influence of some potential sampling and survivorship biases [@brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014], evaluation of which is beyond the scope of the current analysis.
At a minimum, our approach provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig. `r size`).
For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
Indeed, in a comparative analysis of methods to detect growth trends, approaches similar to that employed here (regional curve standardization) performed best at growth trend detection [@peters_detecting_2015], supporting the approach used here as current best practice *[but check @hember_tree_2019].*
Yet, the trends detected here remain potentially subject to a variety of sampling and survivorship biases that can influence overall trends in positive or negative directions (*REFS*).

@@

Changes over time in growth rate are attributable to directionally changing environmental conditions, consisting of some combination of changes in the biotic neighborhood (e.g., competition, crown position) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
The majority of negative growth trends observed here, *to the extent that they are robust to potential sampling and survivorship biases*, are probably attributable to increasing competition as secondary stands age. 
Several of our sites (**HKK, SCBI, LDW, HF, ...** *Coauthors, I need help here.*) are known to have suffered major disturbances within or shortly prior to the analysis time frame [Appendix `r site_details_appendix`; REFS: @baker_disturbance_2005; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
*Indeed, stand succession has been recognized as a driver of growth declines in tree ring records... (REFS).*
Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.
For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.
